---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ratatouille

<!-- badges: start -->

[![R-CMD-check](https://github.com/inbo/ratatouille/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/inbo/ratatouille/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/inbo/ratatouille/graph/badge.svg)](https://app.codecov.io/gh/inbo/ratatouille)

<!-- badges: end -->

## Rationale

This repository contains the functionality to access the unprocessed management data of [RATO vzw](https://oost-vlaanderen.be/wonen-en-leven/natuur-en-milieu/overlastsoorten/rattenbestrijding-.html) and the province of West Flanders via an API call to an [ESRI ArcGIS instance](https://developers.arcgis.com/documentation/). Credentials are required to access this data. These credentials are stored as a github secrets under the `RATO_USER`, `RATO_PWD` and `WFL_USER`, `WFL_PWD`environmental parameters respectively, and can be set locally by editing the `.Renviron` file.

## Set Up your credentials locally

The easiest way to store the required username and password necessary to access a data source is by storing them in your `.Renviron` file. An easy way to edit this file is by using a function from [the usethis package](https://usethis.r-lib.org/).

```{r, eval = FALSE}
# install.packages("usethis")
usethis::edit_r_environ()
```

The contents of the file may contain other secrets or settings used by other scripts or packages, but can be edited to have lines like:

``` bash
RATO_USER = "your username"
RATO_PWD = "your password"
```

After restarting your R instance you should be able to fetch data from the RATO database without being prompted for your credentials.

## Installation

You can install the development version of ratatouille from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("pak")
pak::pak("inbo/ratatouille")
```

## Example

Once you have your credentials stored, getting raw data from RATO is as easy as:

```{r, example of main function, message = FALSE}
library(ratatouille)
rato_data <- ratatouille(source = "rato")
```

The resulting object looks like this (but has many more columns):

```{r, slice raw output, echo = FALSE}
dplyr::slice_sample(rato_data, n = 5) |>
  dplyr::select(dplyr::all_of(
    c(
      "Soort",
      "Materiaal_Vast",
      "Materiaal_Consumptie",
      "Laatst_Bewerkt_Datum"
    )
  )) |> 
  knitr::kable()
```

For this example only a few fields are shown because there is the possibility of personal data in other fields.

### Available columns

These are all columns available for the `rato` data source:

```{r, list columns ,echo = FALSE, message = FALSE}
dplyr::slice(rato_data, 0) |>
  dplyr::glimpse()
```

## Species in the dataset

This is raw data, some cleaning will be required. Dutch names and GBIF taxon ids are provided, but these last are often unchecked and may contain errors. Note that you can only download data for one source at a time.

```{r, count species, message = FALSE}
 purrr::map(c("wfl", "rato"), ~ratatouille(source = .x)) |>
  purrr::list_rbind() |>
  dplyr::count(Soort, sort = TRUE)
```
